---
title: "Chi square distribution"
output: html_document
date: "2024-01-17"
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(shiny)
```

### Chi-square $\chi^2$ distribution

Chi-square $\chi^2$ distributions are a family of continuous probability distributions. They’re widely used in hypothesis tests, including the chi-square goodness of fit test and the chi-square test of independence and categorical analysis.

Very few real-world observations follow a chi-square distribution. The main purpose of chi-square distributions is hypothesis testing, not describing real-world distributions.



```{r, echo=FALSE}
shinyApp(

  ui = fluidPage(
 
            sliderInput("dfs",
                        "Degrees of freedom:",
                        min = 1,
                        max = 20,
                        value = 3),
    plotOutput("distPlot")
  ),

  server = function(input, output) {
    
     
    output$distPlot = renderPlot({
 x <- rchisq(50000, df = input$dfs)
hist(x, 
     freq = FALSE,
     xlim=c(0,50),
     ylim=c(0, 0.3),
     main= paste("chi square df =" , input$dfs) )
curve(dchisq(x,df= input$dfs),from=0, to=50, n=5000,col='red',lwd=2, add=T)
    }

)
  },

  options = list(height = 500)

)
```



### Relationship to the standard normal distribution

Chi-square distributions are useful for hypothesis testing because of their close relationship to the standard normal distribution. 

We will use the standard normal distribution to derive the pdf for chi square distribution with one degree of freedom. 

Let random variable Y be defined as $Y = X^2$ where $X$ has normal distribution with mean 0 and variance 1 (that is $X \sim N(0, 1)$).
$f_X(x) = \frac{1}{\sqrt{2 \pi}} e^{-\frac{x^2}{2}}$

Then,\

For $y < 0, f_Y(y) = P(Y < y) = 0$ and 

For $y \ge 0$, 

\begin{align*}
 f_Y(y) &= P(Y < y) = P(X^2 < y) = P(|X| < y) = P(-\sqrt{y} < X < \sqrt{y}) \\
 & = F_X(\sqrt{y}) - F_X(-\sqrt{y}) =  F_X(\sqrt{y}) - (1-F_X(\sqrt{y})) = 2 F_X(\sqrt{y}) -1  \\
f_Y(y) &= \frac{d}{d y} F_Y(y) = 2 \frac{d}{d y} F_X(y) -0 = 2 \displaystyle \frac{d}{d y} \int_{-\infty}^{\sqrt{y}} \frac{1}{\sqrt{2 \pi}} e^{-\frac{t^2}{2}} dt \\
&= 2 \frac{1}{\sqrt{2 \pi}} e^{-\frac{y}{2}} \sqrt{y}_y^{'} = 2 \frac{1}{\sqrt{2 \pi}} e^{-\frac{y}{2}} \frac{1}{2} y^{-\frac{1}{2}}\\ 
&= \frac{1}{\sqrt{2} \Gamma(\frac{1}{2})} y^{-\frac{1}{2}} e^{-\frac{y}{2}}
\end{align*}


Note:

For positive integer arguments, the gamma function coincides with the factorial. That is,  
  
  $\Gamma(n) = (n-1)!$    
  
  and hence
\begin{align*}
 \Gamma(1) & = 1 \\
 \Gamma(2) & = 1  \\
 \Gamma(3) & = 2 \\
 \Gamma(4) & = 6 \\ 
 \Gamma(5) & = 24
\end{align*}

and so on. For non-positive integers, the gamma function is not defined.

A well-known formula, which is often used in probability theory and statistics, is the following:

$\Gamma(\frac{1}{2}) = \sqrt{\pi}$, we will use Gamma function $ \Gamma(z)$ integral to prove:

\begin{align*}
 \Gamma(z) & = \int_0^{\infty} x^{z-1}  exp(-x) dx \\
 \Gamma(\frac{1}{2}) & = \int_0^{\infty} x^{-\frac{1}{2}}  exp(-x) dx \\
 &= 2 \int_0^{\infty}  exp(-t^2) dt, \quad \text{change of variable} t= x^{1/2} \\
\end{align*}

 
## Alternative proof directly using the change of variable formula

The change of variable formula is used very often in deriving pdf. We have used it the above proof, and for a monotone transformation $y = g(x)$ the general formula is as following:

\begin{align*}
 f_Y(y) & = \sum_i f_X(g_i^{-1}(y)) \left | \frac{dg_i^{-1}(y)}{dy} \right |
\end{align*}

 In this case the change is not monotonic, because every value of $Y$ has two corresponding values of $X$ (one positive and negative). However, because of symmetry, both halves will transform identically, i.e.
 
 \begin{align*}
 f_Y(y) & = 2 f_X(g^{-1}(y)) \left | \frac{dg^{-1}(y)}{dy} \right |
\end{align*}

In this case, the transformation is: $x= g^{-1}(y) = \sqrt{y}$, and its derivative is $\frac{dg^{-1}(y)}{dy} = \frac{1}{2 \sqrt{y}}$.

So here,

 \begin{align*}
 f_Y(y) & = 2 \frac{1}{ \sqrt{2 \pi}} e^{-y/2} \frac{1}{ \sqrt{2 y}} = \frac{1}{ \sqrt{2 \pi y}} e^{-y/2}
\end{align*}

And one gets the chi-squared distribution, noting the property of the gamma function: $\Gamma(1/2) = \sqrt{\pi}$



